---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
We are going to assume that the dataset is in the working folder. Step 1 will be to load the data into R and see a quick preview. 

```{r}
activitydata <- read.csv('activity.csv')
head(activitydata)
```

Lets take a look at the structure of the dataset.
```{r}
str(activitydata)
```

We see that date is stored as a factor. Lets convert it to the correct format.
```{r}
activitydata$date <- as.Date(activitydata$date)
str(activitydata)
```

The date now is correctly formatted. Let's proceed to answer the questions.


## What is mean total number of steps taken per day?
**Part 1**

The goal is to make a histogram of the total steps taken per day. Using the dplyr package we can group and sum the activity data by date. Then we can plot a histogram with that data. 
```{r}
library(dplyr)
activitydata_date <- group_by(activitydata,date)
stepsperday<-summarise(activitydata_date,total=sum(steps))
stepsperday
```

Now we plot the histogram.
```{r}
hist(stepsperday$total,main="Histogram of steps per day",xlab="Total steps")
```

**Part 2**

The mean and median steps per day can be easily calculated . Ignore NA to get the correct responses. The mean is :
```{r}
mean(stepsperday$total,na.rm = TRUE)
```

The median is:
```{r}
median(stepsperday$total,na.rm = TRUE)
```


## What is the average daily activity pattern?

**Part 1**

We need a time series depicting average activity (y axis) by the time interval (x axis). To do this we can group the data by time interval and summaraise it by mean. 

```{r}
library(dplyr)
activitydata_interval <- group_by(activitydata,interval)
stepsperinterval<-summarise(activitydata_interval,average=mean(steps,na.rm = TRUE))
stepsperinterval
```

Now we can plot the time series using ggplot. 

```{r}
library(ggplot2)
g<-ggplot(stepsperinterval,aes(interval,average))
g + geom_point() + geom_line(group='1')
```

**Part 2**

Based on this we observe that the interval with the maximum average steps is somewhere in the 800's . Lets find out the exact interval using which.max.

```{r}
maxinterval <- stepsperinterval[which.max(stepsperinterval$average),]
maxinterval
```

So we can see that it is `r maxinterval[1,1]` 



## Imputing missing values
**Part 1**

We need to find the total number of missing values in the dataset. This can be seen using the summary function. 

```{r}
summary(activitydata)
```

We can see that there are 2304 NA's in the steps column. 

**Part 2**

The strategy we use to replace these values is to calculate the mean for that interval and use it to replace the missing value. 

The idea behind this is that on an average a person is expected to maintain a similar activity level throughout the day. This will let us replicate that. 

We first right a function to replace an NA value with a mean value.

```{r}
impute.mean<-function(x) replace(x,is.na(x),mean(x,na.rm=TRUE))
```

**Part 3**

We can now replicate the original dataset and use dplyr to apply this function to all the records

```{r}
activitydata_imputed <- activitydata %>% group_by(interval) %>% mutate(steps=impute.mean(steps))
head(activitydata_imputed)
```

As we can observe, the previously missing values have been replaced by the interval averages. 

**Part 4**

We can now recreate the histogram for the new dataset.
```{r}
activitydata_imputed_date <-group_by(activitydata_imputed,date)
stepsperday_imputed<-summarise(activitydata_imputed_date,total=sum(steps))
hist(stepsperday_imputed$total,main="Histogram of steps per day",xlab="Total steps")
```

Notice that compared to the original histogram the frequencies are much higher in the new one. 

Let's see the mean and median for the imputed dataset. Notice that na.rm is not needed this time as there are no NA values in the dataset.

The mean is:
```{r}
mean(stepsperday_imputed$total)
```

The median is:
```{r}
median(stepsperday_imputed$total)
```

The mean remains the same but the median is now equal to the mean in the imputed dataset.


## Are there differences in activity patterns between weekdays and weekends?
We can use a convenient function called is.weekend in the chron package along with the mutate function to add a weekday,weekend column. 

```{r}
library(chron)
activitydata_imputed<- activitydata_imputed %>% mutate(daytype=ifelse(is.weekend(date),'weekend','weekday'))
```

We then convert the daytype column to a factor and verify the results.

```{r}
activitydata_imputed$daytype<-as.factor(activitydata_imputed$daytype)
str(activitydata_imputed)
head(activitydata_imputed)
```

**Part 2**

We can now create a panel time series plot to see differences in activity across weekdays and weekends. 

```{r}

activitydata_imputed_interval <- group_by(activitydata_imputed,daytype,interval)
stepsperinterval_imputed<-summarise(activitydata_imputed_interval,average=mean(steps,na.rm = TRUE))
g<-ggplot(stepsperinterval_imputed,aes(interval,average))
g + facet_grid(daytype ~.) + geom_point() + geom_line(group='1')
```

